Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-02-23 paper/paper1/ms.tex
The introduction is clean now

\documentclass[modern]{aastex631}
\bibliographystyle{aasjournal}
\turnoffedit
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{censor}
\usepackage{mathtools}
\usepackage{outlines}
\usepackage{amsmath,bm}

@@

The \texttt{blas\'e} framework aims to take in an observed astronomical spectrum and report back an extensible semi-empirical spectral forward model, replete with the postition, amplitude, width, and shape of all of its spectral lines, automatically.  Solving this problem is hard, for at least four reasons.  First, the spectral lines may overlap, and so the assignment of one line may be partially degenerate with the assignment of some other adjacent line.  Second, extremely wide line wings blend into the continuum, such that the placement of the continuum level may become ill-defined.  Third, the extent of line blending and realized line shape depends strongly on the spectral resolution of the spectrograph, the rotational broadening of the star, and possibly the instrumental configuration at the time of observation.  Finally, telluric absorption lines commingle with the astronomical spectral lines of interest, censoring some spectral regions entirely, or partially confounding other lines with chance allignments.

@@

But many new and important questions in the fields of stars and exoplanets aspire to reach the margins of what the entire dataset can inform.  In particular, data from \emph{high-grasp} \'echelle spectrographs possess simultaneously high spectral resolving power and high bandwidth, yielding tens of thousands or possibly millions of independent spectral data points for each star, substar, or exoplanet.  Those applications that seek to gain signal by ``stacking'' or cross-correlating spectral lines can hypothetically gain huge boosts in the accessible signal-to-noise ratio compared to a single or few lines.  Most manual and semi-automated methods cannot take advantage of the entire spectral bandwidth, and may fail to achieve the hypothetical promise of these high-bandwidth spectrographs.

For example, exoplanet cross-correlation spectroscopy \citep{2013MNRAS.436L..35B} hinges on accurate molecular spectral templates to detect and characterize the atmospheres of exoplanets.  Imperfections in these templates can mute the perceived signal strength of these atmospheric features \citep{2015A&A...575A..20H}.  
In extreme precision radial velocity (EPRV) applications, cross-correlation methods work, but addressing telluric absorption at the $\mathrm{cm/s}$ level may require joint modeling of the star and the Earth's atmospheric absorption before convolution with an instrumental kernel\footnote{\url{https://speakerdeck.com/dwhgg/telluric-line-hack-week-wrap-up}}.  
In the case of Doppler imaging, an accurate underlying template is needed to detect longitudinally symmetric structures such as polar spots \citep{1983PASP...95..565V, roettenbacher16} or zonal bands \citep{Crossfield14,2021ApJ...906...64A}.  The \texttt{blas\'e} framework was developed as an enabling technology for these and many other applications for which an accurate \emph{evaluable} noise-free template is desired, or for which the properties of a large number of individual spectral lines is sought.

Existing frameworks have already overcome aspects of these challenges, or have been purpose-built for specific applications.  The \texttt{specmatch} synthetic template matching tool produces noise-free nearest neighbor templates given an input spectrum \citep{2015PhDT........82P}.  Several practical barriers limit the accuracy of using precomputed synthetic spectral models alone. First and foremost, real stars are usually more complicated than our simplified models of them. Real spectra often vary over more dimensions that our models do.  Conspicuous examples of these hidden variables can be found in protostars: starspots, accretion veiling, dust extinction, and magnetic Zeeman splitting. Jointly modeling all of these phenomena alongside the intrinsic stellar photosphere is challenging.

\subsection{This paper: towards semi-empirical models}
\begin{itemize}
 \item So in practice, tradeoffs between model flexibility and physical self-consistency
 \item Semi-empirical models may be a middle ground: informed from models, but revised with data
 \item The radiative transfer step is computationally expensive, so we want to avoid redoing that
 \item Instead focus on pre-computed models that have that wisdom baked-in
 \item We are essentially compressing pre-computed models into evaluable, interpretable models
 \item Transfer learning is one plausible approach: pre-train on pre-computed models, transfer to data
 \item Neural network frameworks make this possible with autodiff
 \item We use the autodiff and GPU acceleration, but not the neural architectures
 \item We do not use the neural architectures because we have a good understand for how a spectrum is generated
 \item Instead of tuning weights of a neural network, we tune the atomic and molecular properties of a spectrum
\end{itemize}

\section{Methodology}

@@

>>>>

\documentclass[modern]{aastex631}
\bibliographystyle{aasjournal}
\turnoffedit
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{censor}
\usepackage{mathtools}
\usepackage{outlines}
\usepackage{amsmath,bm}
\usepackage{hyperref}
\Urlmuskip=0mu  plus 10mu

@@

The metaphorical holy grail of astronomical spectroscopy is a function that takes in an observed astronomical spectrum and reports back the postition, amplitude, width, and shape of all of its spectral lines, automatically, accurately, and precisely.  The function would go further, to report back the systemic radial velocity (RV) and rotational broadening ($v\sin{i}$).  Finally the function would---in an \emph{Inception}-like manner---provide what it believes to be the interpretable generating function that produced the data, so that we may gain insights on future examples of this or other stars.  Solving this problem is hard, for at least four reasons.  First, the spectral lines may overlap, and so the assignment of one line may be partially degenerate with the assignment of some other adjacent line.  Second, extremely wide line wings blend into the continuum, such that the placement of the continuum level may become ill-defined.  Third, the extent of line blending and realized line shape depends strongly on the spectral resolution of the spectrograph, the rotational broadening of the star, and possibly the instrumental configuration at the time of observation.  Finally, telluric absorption lines commingle with the astronomical spectral lines of interest, censoring some spectral regions entirely, or partially confounding other lines with chance allignments.

@@

But many new and important questions in the fields of stars and exoplanets aspire to reach the margins of what the entire dataset can inform.  In particular, data from high-\emph{grasp} \'echelle spectrographs possess simultaneously high spectral resolving power and high bandwidth, yielding tens of thousands or possibly millions of independent spectral resolution elements for each star, substar, or exoplanet.  Those applications that seek to gain signal by ``stacking'' spectral lines or cross-correlating templates can hypothetically gain huge boosts in the accessible signal-to-noise ratio compared to a single or few lines.  Most manual and semi-automated methods cannot take advantage of the entire spectral bandwidth, or rely on exact knowledge of the underlying templates and may fail to achieve the hypothetical promise of these high-bandwidth spectrographs \citep{2020AJ....160..198H}.

For example, exoplanet cross-correlation spectroscopy \citep{2013MNRAS.436L..35B} hinges on accurate molecular spectral templates to detect and characterize the atmospheres of exoplanets.  Imperfections in these templates can mute the perceived signal strength of these atmospheric features \citep{2015A&A...575A..20H}.  
In extreme precision radial velocity (EPRV) applications, cross-correlation methods work, but addressing telluric absorption at the $\mathrm{cm/s}$ level may require joint modeling of the star and the Earth's atmospheric absorption before convolution with an instrumental kernel\footnote{\url{https://speakerdeck.com/dwhgg/telluric-line-hack-week-wrap-up}}.  
In the case of Doppler imaging, an accurate underlying template is needed to detect longitudinally symmetric structures such as polar spots \citep{1983PASP...95..565V, roettenbacher16} or zonal bands \citep{Crossfield14,2021ApJ...906...64A}.

Existing open-source frameworks have overcome some of these challenges, or have been purpose-built for specialized applications. 
These frameworks include \texttt{ROBOSPECT} \citep{2013PASP..125.1164W}, \texttt{specmatch} \citep{2015PhDT........82P}, \texttt{specmatch-emp} \citep{2017ApJ...836...77Y}, \texttt{wobble} \citep{2019AJ....158..164B}, \texttt{starfish} \citep{czekala15}, \texttt{sick} \citep{2016ApJS..223....8C}, \texttt{psoap} \citep{2017ApJ...840...49C}, \texttt{FAL}\footnote{\url{https://github.com/pacargile/FAL}}, CHIMERA \citep{2015ApJ...807..183L}, the \texttt{Cannon} \citep{2017ApJ...836....5H},  \texttt{MINESweeper} \citep{2020ApJ...900...28C}, and recently \texttt{ExoJAX} \citep{2022ApJS..258...31K}.  
The design of these frameworks necessarily have to make a choice in the bias-variance tradeoff: is the tool more \emph{data}-driven or more \emph{model}-driven?  The statistical tradeoff can be viewed as a concession in physical self-consistency for model flexibility: more or fewer parameters.

A key new enabling technology breaks the historic limits of data-model comparisons for astronomical spectroscopy.  Automatic differentiation \citep[``autodiff'',][]{2015arXiv150205767G} and the backpropagation algorithm has revolutionized machine learning and neural network architecture design, and is increasingly applied in astrophysical data analysis contexts, \emph{e.g.} kernel phase coronography with \texttt{poppy} \citep{2021ApJ...907...40P}, and exoplanet orbit fitting with \texttt{exoplanet} \citep{2021JOSS....6.3285F}.  Of the spectroscopy frameworks mentioned above, \texttt{wobble} \citep{2019AJ....158..164B} and \texttt{ExoJAX} \citep{2022ApJS..258...31K} employ autodiff technology.  \texttt{wobble} treats each pixel as a tunable control point, producing $\sim10^5$ parameters for a modern stellar spectrum.  The \texttt{ExoJAX} framework has only $\sim$dozens of tunable parameters that describe the fundamental physical properties controlling a brown dwarf atmosphere.  These two autodiff-aware frameworks span the extreme ends of non-parametric and parametric modeling for spectroscopy.

In this paper we show that autodiff-aware semi-empirical models offer a cozy middle ground: informed from self-consistent models, but refined with data.  In Section \ref{methodology} we show how to supplant the computationally expensive radiative transfer, by cloning pre-computed synthetic spectral models.  In Section \censor{XX} we describe how to fine-adjust the cloned models to data in a transfer-learning step.  Section \censor{XX} exhibits some injection-recovery type tests on simulated data of known properties.  In Section \censor{XX} we show real-world data from high-grasp near-IR spectrographs.  Finally we discuss many conceivable extensions to the framework, and directions for future work in Section \censor{xx}.

\section{Methodology}\label{methodology}

@@

\subsection{Comparison to existing frameworks}

The \texttt{specmatch} synthetic template matching tool produces noise-free nearest neighbor templates given an input spectrum \citep{2015PhDT........82P}.  Several practical barriers limit the accuracy of using precomputed synthetic spectral models alone. First and foremost, real stars are usually more complicated than our simplified models of them. Real spectra often vary over more dimensions that our models do.  Conspicuous examples of these hidden variables can be found in protostars: starspots, accretion veiling, dust extinction, and magnetic Zeeman splitting. Jointly modeling all of these phenomena alongside the intrinsic stellar photosphere is challenging.
